---
title: "From a word to its sounds"
subtitle: "Griffin & Ferreira (2006), sections 1.3 to 1.5"
author: "Utku Turk"
institute: "LING 449U, How words are built"
date: "2026-09-08"
description: "Function words and morphemes, why the sounds of a word are rebuilt every time you say it, and how discrete the stages really are. Properties 9 to 14."
categories: [Week 2]
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: c/t
    incremental: false
    history: false
    transition: none
    fig-align: center
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)

theme_set(
  theme_minimal(base_size = 18) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "none",
      plot.subtitle = element_text(colour = "grey30", size = 14)
    )
)

help_col <- "#2166AC"
hurt_col <- "#B2182B"
null_col <- "grey55"
```

## Where we got to {.smaller}

1. **Two stages.** Wrong word said perfectly vs. right word scrambled.
2. **A family activates.** *van* wakes up *bus*.
3. **They compete.** Two good names (*TV*, *television*) makes you slower.
4. **Class fences the competition.** Nouns replace nouns.
5. **Meaning properties matter.** *vampire* beats *spirit*, and gets replaced more often.
6. **Intention to name.** An ignored banana does not interfere; a written *banana* does.
7. **Selection leaves a trace.** Name a zebra, then call the giraffe *zebra*.
8. **It can fail halfway.** TOT: gender and syllable count, no form.

## The one result to keep

::: {.soa}
::: {.soa-track}
::: {.soa-pic}
picture
:::
::: {.soa-word}
distractor
:::
:::
::: {.soa-axis}
[-400 ms]{.t1} [-200]{.t2} [0]{.t3} [+200]{.t4} [+400]{.t5}
:::
:::

- **Early** distractor, **semantic** relation: interference. Selection is running.
- **Late** distractor, **phonological** relation: facilitation. Sounds are running.

::: {.fragment .claim}
Same picture, same distractor, different SOA, opposite sign. The timing is the
evidence for the stages.
:::

::: footer
Schriefers, Meyer & Levelt 1990
:::

## Today

- 1.3, function words and morphemes (9)
- 1.4, assembling the sounds (10 to 13)
- 1.5, time course (14 and 15)

::: {.fragment}
Property 9 is the only place in the whole chapter where inflection gets a
paragraph. We are going to sit on it for a while.
:::

## 1.3 Function words and morphemes

::: {.claim}
Some words and morphemes are not selected by meaning. They are selected by the
words around them.
:::

- English: *some pasta* vs. *a noodle*, count or mass.
- Swedish: *ett bord* vs. *en stol*, gender.
- French: *mon chapeau*, *ma table*, but *mon ampoule*, gender **and** phonology.

::: footer
Property 9. Janssen & Caramazza 2003
:::

## The evidence

- **Gender interference.** Dutch *het groene huis* is slowed by the distractor
  *tafel* (de) more than by *been* (het).
- **Agrammatism.** Function words are impaired even though they are frequent and
  phonologically simple.
- **Speech errors.** Function words slip differently from content words.

::: footer
Schriefers 1993; Bates, Wulfeck & MacWhinney 1991; Garrett 1975; Dell 1990
:::

## Two ways to get a function word {.smaller}

:::: {.columns}
::: {.column width="50%"}
**Positional frames** (Garrett)

Function words are *part of the frame* content words get slotted into.

Retrieved by a different route than nouns, but sensitive to meaning.
:::
::: {.column width="50%"}
**Indirect election** (Levelt, Ferreira)

Retrieve the noun first. The noun's syntactic knowledge then triggers the
determiner.

Word first, then its dependents.
:::
::::

::: {.fragment .claim}
Both deny direct meaning-to-function-word mapping. They disagree on whether the
frame exists before the content word does.
:::

## Property 9: what this is doing to us

::: {.prompt}
- The chapter's split is content vs. function. Where does *-ed* go? It is not a
  word, and it is not free.
- If the noun triggers the determiner, what triggers plural *-s* on that noun?
- Is a positional frame a syntactic structure, or a phonological one?
:::

## 1.4 The sounds are built anew

- Not retrieved as intact wholes. Assembled every single time.
- *hand* alone vs. *hand it* = "han-dit". The /d/ resyllabifies.
- Repeating *kem-til-fer* is faster than *kem-til-ner*: shared CVC structure,
  different content.

::: {.fragment .claim}
The stored thing is not a pronunciation. It is a set of segments plus a metrical
frame, put together on the spot.
:::

::: footer
Property 10. Levelt et al. 1999; Sevald, Dell & Cole 1995
:::

## Property 11: experience shapes the errors {.smaller}

- **Lexical bias.** Slips make real words more often than chance.
- **Position constraints.** New sequences still respect syllable position.
- **Least predictable slips.** Word-initial /b/ in *bicycle* separates more
  easily than the /s/ and /k/. Pig Latin, *helter skelter*.

::: {.fragment .claim}
So which consonant slips should depend on which language you speak.
:::

::: footer
Baars, Motley & MacKay 1975; Dell & Reich 1981; Shattuck-Hufnagel 1987
:::

## The same slip, two languages {.smaller}

```{r}
#| label: crosslinguistic-slips
#| echo: false
#| fig-width: 9
#| fig-height: 3.4

berg <- data.frame(
  lang = rep(c("Germanic", "Spanish"), each = 2),
  pos = rep(c("Word-initial", "Other syllable-initial"), 2),
  share = c(68, 32, 51, 49)
)
berg$pos <- factor(berg$pos, levels = c("Word-initial", "Other syllable-initial"))

ggplot(berg, aes(lang, share, fill = pos)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey45") +
  scale_fill_manual(values = c("Word-initial" = hurt_col,
                               "Other syllable-initial" = null_col)) +
  coord_flip() +
  labs(
    x = NULL, y = "Share of consonant slips (%)", fill = NULL,
    subtitle = "Schematic. Dashed line = no asymmetry."
  ) +
  theme(legend.position = "top")
```

Spanish does not distinguish word-initial from other syllable-initial
consonants, and Spanish slips do not either.

::: footer
Berg 1991, 1998; Shattuck-Hufnagel 1987
:::

## Property 11: two timescales

- **Short term.** If /f/ only appears syllable-initially inside one experiment,
  slips involving /f/ go syllable-initial within an hour.
- **Long term.** Frequency and age of acquisition: common words come out up to
  **100 ms** faster.

::: {.fragment}
So where in the system does frequency live?
:::

::: footer
Dell, Reed, Adams & Meyer 2000; Oldfield & Wingfield 1964
:::

## Where frequency bites {.smaller}

```{r}
#| label: frequency-locus
#| echo: false
#| fig-width: 9
#| fig-height: 3.4

freq <- data.frame(
  outcome = c(
    "Phonological substitutions",
    "Tip-of-the-tongue states",
    "Semantic substitutions,\nunder time pressure",
    "Semantic substitutions,\nno time pressure"
  ),
  penalty = c(1.00, 0.85, 0.45, 0.05)
)
freq$outcome <- factor(freq$outcome, levels = rev(freq$outcome))
freq$stage <- c("form", "form", "selection", "selection")

ggplot(freq, aes(penalty, outcome, fill = stage)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c(form = hurt_col, selection = null_col)) +
  scale_x_continuous(limits = c(0, 1.1), breaks = NULL) +
  labs(
    x = "How much a low-frequency word costs you",
    y = NULL,
    subtitle = "Schematic. Relative sizes only; the point is which bar is near zero."
  )
```

::: {.claim}
Frequency looks like a **phonological encoding** effect, not a selection effect.
:::

::: footer
Dell 1990; Stemberger & MacWhinney 1986; Vitkovitch & Humphreys 1991
:::

## Property 11: homophones {.smaller}

*ball* the object and *ball* the dance: different meanings, same sounds.

::: {.fragment}
- The rare *ball* slips less than equally rare words.
- Practicing one meaning helps an aphasic produce the other, for a week.
- *cherry pit* resolves a TOT for *Brad Pitt*.
:::

::: {.fragment .claim}
Shared phonology, separate lexical entries. This is the cleanest case in the
chapter of one form serving two meanings, and the chapter never asks what
happens when one *meaning* has two forms.
:::

::: footer
Dell 1990; Biedermann, Blanken & Nickels 2002; Burke et al. 2004
:::

## Property 12: sounds go left to right

The manipulation: you are told in advance that every word in this block starts
the same way. Or ends the same way. Does it help?

::: {.fragment}
- **Initial** overlap (*tile* for *tiger*) acts at earlier SOAs than **final**
  overlap (*liar*).
- The advance-knowledge benefit needs shared onset **and** shared metrical
  structure.
:::

::: footer
Meyer & Schriefers 1991; Meyer 1990, 1991; Roelofs & Meyer 1998
:::

## What foreknowledge buys you


```{r}
#| label: prep-benefit
#| echo: false
#| fig-width: 9
#| fig-height: 3.6

prep <- data.frame(
  what = c(
    "Shared onset,\nsame metrical frame",
    "Shared onset,\nframe varies",
    "Shared final sounds"
  ),
  benefit = c(35, 5, -1)
)
prep$what <- factor(prep$what, levels = rev(prep$what))
prep$sign <- ifelse(prep$benefit > 15, "help", "none")

ggplot(prep, aes(benefit, what, fill = sign)) +
  geom_col(width = 0.6) +
  geom_vline(xintercept = 0, linewidth = 0.6) +
  scale_fill_manual(values = c(help = help_col, none = null_col)) +
  scale_x_continuous(limits = c(-10, 45)) +
  labs(
    x = "Speed-up vs. an unrelated block (ms)",
    y = NULL,
    subtitle = "Schematic magnitudes, real directions."
  )
```

::: {.claim}
You can prepare a word prefix. You cannot prepare a word suffix.
:::

::: footer
Meyer 1990, 1991; Roelofs & Meyer 1998
:::

## Property 12: the problem for morphology

::: {.prompt}
Suffixes are at the end. If the end is exactly the part you cannot prepare in
advance, how is *-ed* ever ready on time?
:::

::: {.fragment .aside}
Also note: a phonological *feature* (place of articulation) gives no benefit.
The unit of preparation is the whole segment.
:::

::: {.fragment}
And length effects mostly vanish once confounds are controlled. Multisyllabic
words are not slower to prepare than monosyllabic ones.
:::

::: footer
Roelofs 1999; Griffin 2003; Bachoud-Levi et al. 1998
:::

## Property 13: similar sounds, no fixed sign

```{r}
#| label: sign-flip
#| echo: false
#| fig-width: 10
#| fig-height: 4.4

flip <- data.frame(
  task = c(
    "Related distractor (liar / lion)",
    "Dense phonological neighbourhood",
    "Previous trial shares final sounds",
    "Previous trial shares onset (hond / hoed)",
    "Repeating pairs with shared onsets"
  ),
  effect = c(-25, -18, -12, 16, 30)
)
flip$task <- factor(flip$task, levels = rev(flip$task))
flip$sign <- ifelse(flip$effect < 0, "helps", "hurts")

ggplot(flip, aes(effect, task, colour = sign)) +
  geom_vline(xintercept = 0, linewidth = 0.6) +
  geom_segment(aes(x = 0, xend = effect, yend = task), linewidth = 1.6) +
  geom_point(size = 4.5) +
  scale_colour_manual(values = c(helps = help_col, hurts = hurt_col)) +
  scale_x_continuous(limits = c(-35, 40)) +
  labs(
    x = "Effect of phonological similarity (ms)",
    y = NULL,
    subtitle = "Left of zero helps, right of zero hurts. Schematic magnitudes."
  )
```

::: footer
Vitevitch 1997, 2002; Schriefers et al. 1990; Sevald & Dell 1994; Wheeldon 2003
:::

## Property 13: why the sign flips

::: {.claim}
Semantically related words **compete**. Phonologically related words **support**
each other, except when the task makes you sequence them.
:::

::: {.fragment}
- Retrieval of segments: support.
- Sequencing segments: competition. Finishing /i/ with /k/ makes finishing /i/
  with /p/ harder.
:::

::: {.fragment}
So *balaclava* for *baklava* is probably not competition. It is a **near miss**:
incomplete retrieval, defaulting to a similar stored form.
:::

::: footer
Dell & O'Seaghdha 1992; Burke et al. 1991
:::

## 1.5 How discrete are the stages? {.smaller}

:::: {.columns}
::: {.column width="50%"}
**Strictly staged**

Sound information is accessed only after selection finishes.

WEAVER++.
:::
::: {.column width="50%"}
**Cascading / interactive**

Unselected candidates leak activation into sound assembly, and sounds feed back.

Dell 1986.
:::
::::

::: {.fragment}
- **Cascading:** preparing to say *couch*, you are faster to read *soda*, because
  the losing synonym *sofa* activated its sounds.
- **Feedback:** semantic substitutions are phonologically more similar to their
  targets than chance. *cat* to *rat*, not *cat* to *dog*.
:::

::: footer
Property 14. Peterson & Savoy 1998; Levelt et al. 1999; Dell & Reich 1981
:::

## Property 14: the discrete reply

::: {.claim}
Mixed errors are not produced more often. They are **caught** less often.
:::

An editor checks the output. The more the intruder resembles the target, the
more likely it slips past.

::: {.fragment .prompt}
Both accounts predict the same distribution of errors you observe. What
experiment separates "generated more" from "detected less"?
:::

::: footer
Motley, Camden & Baars 1982; Levelt 1989; Rapp & Goldrick 2000
:::

## Property 15: message wide, sounds narrow

```{r}
#| label: exchange-distance
#| echo: false
#| fig-width: 8
#| fig-height: 2.6

ex <- data.frame(
  unit = factor(c("Sounds", "Words and morphemes"),
                levels = c("Words and morphemes", "Sounds")),
  syll = c(2.1, 4.1)
)

ggplot(ex, aes(syll, unit)) +
  geom_col(width = 0.55, fill = help_col) +
  geom_text(aes(label = syll), hjust = -0.35, size = 6, colour = "grey25") +
  scale_x_continuous(limits = c(0, 5)) +
  labs(
    x = "Syllables between the two units that exchanged",
    y = NULL,
    subtitle = "Nooteboom 1973. These two numbers are from the chapter."
  )
```

::: {.fragment .claim}
Abstract properties of words are settled further in advance than their sounds
are.
:::

::: footer
Nooteboom 1973
:::

## What else says the message comes first

- Speakers start long utterances at a **higher pitch** than short ones.
- Complex clauses get more pauses and *um*s before them.
- Agreement errors cross clause boundaries less than they cross words.

::: {.fragment .claim}
You plan a proposition-sized message before you start. You do not plan all its
words.
:::

::: footer
Cooper, Soares & Reagan 1985; Bock & Cutting 1992; Goldman-Eisler 1972
:::

## Property 15: how much do you prepare?

- Nouns after the verb are often selected **while you are articulating** the
  first words.
- Minimum unit looks like one **phonological word**: a content word plus its
  clitics, *[beer's a] [good] [thing]*.
- But: more if it is half a compound, more if it is inside the first noun
  phrase.

::: {.fragment .aside}
Aphasic evidence for phrases: patients with lexical-semantic deficits struggle
with *the long, brown hair* but not *the hair was long and brown*. Patients with
phonological deficits show no such difference.
:::

::: footer
Griffin & Bock 2000; Wheeldon & Lahiri 1997, 2002; Martin & Freedman 2001
:::

## Property 15: and time itself

- Speakers insert optional *that* when the next word is slow to arrive.
- *uh* before short delays, *um* before long ones.

::: {.fragment}
Planning scope is not a fixed constant. It is whatever keeps the mouth busy.
:::

::: footer
Ferreira & Dell 2000; Clark & Fox Tree 2002
:::

## What the chapter leaves out

The authors tell you themselves: they have not covered how speakers produce
morphologically complex words.

::: {.fragment}
So the only treatment of morphemes we get is Property 9, where they are filed
next to articles because both depend on the words around them.
:::

::: {.fragment .claim}
Meanwhile all of 1.4 assumes a word is a stem with sounds laid out in order, and
Property 11 already told us that assumption breaks for Arabic.
:::

## Key takeaways

1. Function words are selected by their neighbours, not by meaning.
2. Sound form is assembled fresh, left to right, from segments plus a metrical
   frame.
3. Experience shapes errors at both timescales, and it is language-specific.
4. Semantic neighbours compete; phonological neighbours mostly help.
5. Message planning runs ahead of sound planning, by about a phonological word.

## Bring questions

::: {.prompt}
- Property 12 says word endings cannot be prepared early. Property 9 says
  inflections depend on the stem. Are these the same fact?
- Where in this architecture would *un-* and *-ness* be stored?
- If phonological neighbours support each other, why do morphological families
  ever compete?
:::

::: footer
Reading questions due 11:59pm the night before
:::
